\vspace{2cm} \Large \bf Seasonal influenza circulation patterns and future projections

### 2015-06-15 manuscript/correspondence/response.tex

\vspace{1cm}

Dear editor,

Please find attached our revised manuscript entitled ```nextflu: Real-time tracking of seasonal influenza virus evolution in humans''.  This is a resubmission of manuscript number BIOINF-2015-0560.

We would like to thank you and the reviewers for their supportive comments. We have addressed all comments in detailed point-by-point response below. In the mean-time, we have received very positive feedback from the community, including requests to set up separate website for ebola virus and MERS-coronovirus evolution. Owing to the modular and flexible structure of augur and auspice, implementation of these pipelines was straightforward, underscoring the usefulness of augur and auspice. We hope that you and the reviewers share this impression.
%We also have implemented the LBI calculation that can be used as a coloring option an is dynamically updated when the date-slider is changed.

@@

{\bf The purpose of nextflu is to integrate as much information as possible to get an overview of the currently circulating viruses. As such, nextflu can help making an informed vaccine strain selection. We included quantities that have been suggested to be predictive for this purpose. However, in addition to choosing the clade that best approximates the future virus population, vaccine strain selection faces many practical constraints (virus/serum availability, cell-culture behavior, etc...). By providing the predictors along with explicit temporal dynamics of clades in the tree, the user can evaluate their relative merits and pick the most promising strain, given other constraints. We believe that continually up-to-date displays of circulating viruses will be of the most use to WHO vaccine strain selection. These displays could help week-to-week in making it clear which viruses should be chosen to become test viruses in HI and other phenotypic assays. This said, we would eventually like to experiment with fully predictive models of virus population dynamics (like the models used by Steinbr\"{u}ck et al. 2014 or \L{}uksza and L\"{a}ssig 2014), however, we believe that this work is beyond the scope of nextflu 1.0.}

3. The use case of the application is not clear to me. Can you give a more explicit example how information retrieval with nextflu could foster selection of the ``next successful clade'', rather than browsing the plain GISAID sequences? Of course, nextflu just visualizes existing data and thus cannot make up something that could not also be found in the database itself, but it would be nice to see how you intend to use nextflu.

{\bf nextflu is meant as a tool to explore the most recent influenza sequences, their geographic distribution, and the dynamics of different variants. We have received enthusiastic feedback from the influenza community, including representatives from all five WHO Collaborating Centers and from the GISAID Initiative.

In response to requests from the community, we have recently extended nextflu to ebola virus and MERS coronavirus, underscoring both the need for tools like nextflu and the versatility of augur and auspice.}

@@

{\bf We agree that breaking out the mathematical details behind the frequency estimation is warranted. We've removed these details from the manuscript and describe them online at \href{http://nextflu.org/methods/}{\texttt{nextflu.org/methods/}}.}

\subsection*{Application}

1. You give information about the workflow of the augur pipeline to process data of the underlying GISAID database. What happens if GISAID is updated. Do you have to manually restart augur? Is it possible to trigger such an update by the user of nextflu? If it is not possible, is the term ``real-time'' really satisfied?

{\bf As of now, we need to update manually, which involves only downloading all recent human sequences from GISAID and running the augur processing script. In any case, the deposition of sequences incurs a greater lag than our update frequency. During the North Hemisphere influenza season, when data deposition is frequent, GISAID and nextflu contain sequences within ~3 weeks of the present day. This is as close to `real-time' as present surveillance allows and we believe labeling nextflu as `real-time' is warranted. Also, as surveillance pipelines continue to improve, this lag will continue to decrease.}

@@

2. You describe the data pruning of the GISAID sequences, which also involves a step of pruning the alignment to various criteria. Is this done automatically? It would be nice to give the user the capability to parameterize some of these criteria prior to visualizing the tree.

{\bf Since building aligning $>2000$ sequences and building a tree requires substantial computation, we can't have the user choose sequence filtering criteria and rerun the pipeline on demand. However, to address this concern, we have invested a substantial of effort to provide multiple temporal `resolutions' displaying different sampling windows. For example, \href{http://nextflu.org/H3N2/1y/}{\texttt{nextflu.org/H3N2/1y/}} displays a detailed view of the last year of viral evolution and consequently deep sampling, while \href{http://nextflu.org/H3N2/12y/}{\texttt{nextflu.org/H3N2/12y/}} displays a coarser view of the last 12 years of viral evolution. Furthermore, users who want to customize sampling can easily clone the GitHub repo and run the pipeline locally.}

@@

3. It is rather complicated to zoom in and out or to scroll in the tree. Would it be possible to add according navigation buttons, rather then resetting the whole display when zooming out a bit?

{\bf With the current set-up, this is not readily achieved. But we have added a dynamic inclusion of strain labels once the zoom level becomes high enough that labels can be discerned.}

4. It is difficult to use the date slider to actually move through time since the refresh is only accomplished when releasing the mouse button. Also slight horizontal movements can cause unwanted jumps. Could you maybe provide two buttons to jump to the neighboring date points? Or a functionality to automatically move through time like in a movie.

{\bf Dynamic update of the tree imposes a heavy computational burden. To address this issue, we now show/hide virus circles as the slider is moved to show sampling time, but `gray out' the virus colors while manipulating the time window. We feel that the current compromise of updating tip circles without recalculating predictors and colors is a good compromise. Depending on the speed of the user's computer, this well be more or less smooth. We have also added more flexibility to choose the focal time interval. Users can now manipulate the upper and lower time cut-off separately.}

### 2015-06-15 manuscript/correspondence/response.tex

\vspace{1cm}

Dear editor,

Please find attached our revised manuscript entitled ```nextflu: Real-time tracking of seasonal influenza virus evolution in humans''.  This is a resubmission of manuscript number BIOINF-2015-0560.

We would like to thank you and the reviewers for their supportive comments. We have addressed all comments in detailed point-by-point response below. In the mean-time, we have received very positive feedback from the community, including requests to set up separate website for ebola virus and MERS-coronovirus evolution. Owing to the modular and flexible structure of augur and auspice, implementation of these pipelines was straightforward. This shows that the nextflu framework can be readily applied to very different pathogens.
%We also have implemented the LBI calculation that can be used as a coloring option an is dynamically updated when the date-slider is changed.

### 2015-06-15 manuscript/correspondence/response.tex

% cite package, to clean up citations in the main text. Do not remove.
\usepackage{cite}
%\renewcommand\citeleft{(}
%\renewcommand\citeright{)}
%\renewcommand\citeform[1]{\textsl{#1}}

% Remove brackets from numbering in list of References
\renewcommand\refname{\large References}
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

\usepackage{authblk}
\renewcommand\Authands{ \& }
\renewcommand\Authfont{\normalsize \bf}
\renewcommand\Affilfont{\small \normalfont}
\makeatletter
\renewcommand\AB@affilsepx{, \protect\Affilfont}
\makeatother

\newcommand{\augur}{\textbf{augur}}
\newcommand{\auspice}{\textbf{auspice}}
\newcommand{\nextflu}{\textbf{nextflu}}

@@

\vspace{1cm}

Dear editor,

Please find attached our revised manuscript entitled ```nextflu: Real-time tracking of seasonal influenza virus evolution in humans''.  This is a resubmission of manuscript number BIOINF-2015-0560.

We would like to thank you and the reviewers for their supportive comments. We have addressed all comments in detailed point-by-point response below. In the mean-time, we have received very positive feedback from the community, including enthusiastic responses from the many involved with the World Health Organization Global Influenza Surveillance and Response System. Additionally, we've received requests to set up separate websites for Ebola virus and MERS-CoV evolution. Owing to the modular and flexible structure of \augur{} and \auspice{}, implementation of these pipelines was straightforward. This shows that the nextflu framework can be readily applied to very different pathogens. However, the work with Ebola and MERS-CoV is still somewhat experimental, while the pipelines in \nextflu{} proper are very robust. We believe \nextflu{} will prove highly useful in continued influenza surveillance.

### 2015-06-15 manuscript/correspondence/response.tex

Dear editor,

Please find attached our revised manuscript entitled ```nextflu: Real-time tracking of seasonal influenza virus evolution in humans''.  This is a resubmission of manuscript number BIOINF-2015-0560.

We would like to thank you and the reviewers for their supportive comments. We have addressed all comments in detailed point-by-point response below. In the mean-time, we have received very positive feedback from the community, including enthusiastic responses from the many involved with the World Health Organization Global Influenza Surveillance and Response System. Additionally, we've received requests to set up separate websites for Ebola virus and MERS-CoV evolution. Owing to the modular and flexible structure of \augur{} and \auspice{}, implementation of these pipelines was straightforward. This shows that the nextflu framework can be readily applied to very different pathogens. However, the work with Ebola and MERS-CoV is still somewhat experimental, while the pipelines in \nextflu{} proper are very robust. We believe \nextflu{} will prove highly useful in continued influenza surveillance.

Additionally, we've made the following specific changes to the program / manuscript:

``1) Prediction of successful clades: you could easily implement your Local Branching Index (LBI), instead of the full likelihood calculation, for fitness calculation.''

We provide LBI for each node in the tree at any point in time. This has previously been shown to be predictive of clade success. However, a full analysis of predictive clade dynamics is beyond version 1.0 of the platform. More work on this is forthcoming. We'd like to do this as properly as possible to keep trust in the platform by public health officials. If we provide explicit clade frequency predictions, we want them to be right.

``2) Try to clarify the example case as suggested''

As reviewer 1 suggested, we've clarified the example case.

``3) Try to update the visualization while sliding through the data''

We've clarified and improved the `date slider' behavior.

``4) Move technical parts to supplementary material''

We've moved technical details to an online methods page at \href{http://nextflu.org/methods/}{\texttt{nextflu.org/methods/}}.

``5) It would be convenient to provide the users some control of how the data is being pruned''

We've invested significant effort to making this happen. Users can now select a `data resolution'. \href{http://nextflu.org/H3N2/1y/}{\texttt{nextflu.org/H3N2/1y/}} displays a detailed view of the last year of viral evolution and consequently uses deeper sampling, while \href{http://nextflu.org/H3N2/12y/}{\texttt{nextflu.org/H3N2/12y/}} displays a coarser view of the last 12 years of viral evolution.

``6) Explain how the tool accepts updateds of the database''

This has been clarified.

### 2015-06-15 manuscript/correspondence/response.tex

We would like to thank you and the reviewers for their supportive comments. We have addressed all comments in detailed point-by-point response below. In the mean-time, we have received very positive feedback from the community, including enthusiastic responses from the many involved with the World Health Organization Global Influenza Surveillance and Response System. Additionally, we've received requests to set up separate websites for Ebola virus and MERS-CoV evolution. Owing to the modular and flexible structure of \augur{} and \auspice{}, implementation of these pipelines was straightforward. This shows that the nextflu framework can be readily applied to very different pathogens. However, the work with Ebola and MERS-CoV is still somewhat experimental, while the pipelines in \nextflu{} proper are very robust. We believe \nextflu{} will prove highly useful in continued influenza surveillance.

Additionally, we've made the following specific changes to the program / manuscript:

``1) Prediction of successful clades: you could easily implement your Local Branching Index (LBI), instead of the full likelihood calculation, for fitness calculation.''

We have implemented the LBI calculation that can be used as a coloring option and is dynamically updated when the date-slider is changed.

@@

``4) Move technical parts to supplementary material''

We've moved technical details to an online methods page at \href{http://nextflu.org/methods/}{\texttt{nextflu.org/methods/}}.

``5) It would be convenient to provide the users some control of how the data is being pruned''

To provide customization of data pruning, users can now select between alternative `data resolutions'. For example, \href{http://nextflu.org/H3N2/1y/}{\texttt{nextflu.org/H3N2/1y/}} displays a detailed view of the last year of viral evolution and consequently uses deeper sampling, while \href{http://nextflu.org/H3N2/12y/}{\texttt{nextflu.org/H3N2/12y/}} displays a coarser view of the last 12 years of viral evolution.

### 2015-09-18 auspice/reports/tex/sep-2015.tex
Port @rneher's edit to TeX.

The continued dominance of 3c2.a is corroborated with a large local branching index and comparatively low mutational load as measured by the number of non-epitope substitutions since the root of the tree. Additionally, we observe no rapidly expanding subclades within 3c2.a, 3c3.a or 3c3.b. Within 3c2.a, the fastest branching patterns still appear at the root of the clade, indicative of no further adaptive events occurring (Fig.~\ref{H3N2_lbi}).

>>>>

The continued dominance of 3c2.a is corroborated with a large local branching index (Fig.~\ref{H3N2_lbi}) and comparatively low mutational load as measured by the number of non-epitope substitutions since the root of the tree. Within 3c2.a, one clade (the uppermost clade with in 3c2.a) is starting to dominate but this clade is not characterized by any amino acid differences within HA. We observe no rapidly expanding subclades within 3c2.a or 3c3.b.
